\begingroup
\let\clearpage\relax
\let\cleardoublepage\relax
\let\cleardoublepage\relax

\phantomsection
\chapter*{Acknowledgments}
\addcontentsline{toc}{chapter}{Acknowledgments}
\chaptermark{\itshape Acknowledgments}

First, I must thank my advisors, Chris Rycroft and Alex Townsend. Chris, thank you for bringing such happiness to everything mathematical. My days were always made better by a piece of British humor, a Mac rumor, or the sight of a colored tee. Your dedication to the growth of numerics at Harvard is noble---thank you for making a space for computational research at SEAS. I am grateful for our summer trips to Berkeley and for your mentorship throughout my graduate career. Alex, getting to work with you has been an honor that wholly transformed my graduate school experience. Your unrelenting optimism and excitement for applied mathematics are inspiring, and your creativity contagious. I am grateful to have had a front-row seat. I look back fondly on our conference trips together as well as my trips to Cornell. Your guidance has truly been a gift and I am lucky to call you a mentor and a trusted friend.

My family has been a pillar of support for me during my time in graduate school. To my mom and dad, Marie and Frank: thank you for always being there for me. Your love has helped me through the toughest times and is an unfathomable source of happiness in my life. To my sister, Marisa, and brother-in-law, Kevin: thanks for the fun memories and for your thoughtful advice.  To Ellen: thank you for your unwavering support and infinite kindness.

My friends have been a constant source of encouragement throughout the last five years. To my bandmates in the \href{https://americansymphonyofsoul.com/}{American Symphony of Soul}: Alek, Alex, Andrew, Erich, Kristen, Jackson, Chris, Willie, Mary---getting to play music with you all is a joy and has kept me sane throughout graduate school. I'm lucky to have you as friends. To Jordan and Shruti: thank you for your innumerable sledges, pranks, and late-night hangs. I'll fondly miss our trivia team. To Jon, Carlos, Mark, Nico, Doug, Mike, and Ben: thank you for your lifelong friendships.

I am fortunate to have had many wonderful colleagues and mentors throughout my time in graduate school. Robert Saye has been an endless source of knowledge on high-order methods and high-performance C++, and has been a wonderful collaborator. I'm lucky to have experienced Nick Hale's MATLAB wizardry and deep expertise in spectral methods firsthand, and working with him has been a pleasure. Nick Trefethen's enthusiasm for numerical analysis and its communication has been infectious and inspires me daily. I've greatly benefitted from friendships and conversations with countless others, including Sheehan Olver, Alex Barnett, Rasmus Tamstorf, Keaton Burns, Heather Wilber, Marc Gilles, Andrew Horning, Ben Zhang, Ricardo Baptista, Thomas Fai, Chen-Hung Wu, Eder Medina, Nick Derr, and Luna Lin.

Finally, to my thesis committee---Chris, Alex, Robert, and Efthimios Kaxiras---thank you for taking time to read this thesis and for your detailed feedback.

My research has been generously funded by the Department of Defense through the National Defense Science \& Engineering Fellowship and by National Science Foundation grants 1645445 and 1818757. This research was also supported in part by the Applied Mathematics Program of the U.S. Department of Energy Office of Advanced Scientific Computing Research under contract number DE-AC02-05CH11231. Some computations used resources of the National Energy Research Scientific Computing Center, a Department of Energy Office of Science User Facility supported by the Office of Science of the U.S. Department of Energy under contract number DE-AC02-05CH11231.

\endgroup

\cleardoublepage
